Compositional equivalences based on Open pNets\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-12-03 2016-Weak-FH-Bisim/WeakBisim.tex
Nouveaux TODOs, et modifs dessins

Fig. \ref{SimpleProt:Spec} (right) shows a graphical view of the pNet of the system \emph{specification}:
the full system should be equivalent (e.g. through weak bisimulation)
to the two processes connected simply through a perfect medium.

Fig. \ref{SimpleProt:Impl} show the pNet model of the protocol implementation. When the $Medium$ detects an error (modeled by a local $\tau$ action), it sends back a $m\_error$ message to the $Sender$. The $Sender$ increments its local error counter $c$, and resends the message (including $c$) to the $Medium$, that will, eventually, transmit $m,c$ to the $Receiver$.

@@

The semantics of open pNets will be defined  as an open automaton. An open
automaton is an automaton where each transition composes transitions of several LTSs with
action of some holes, the transition occurs if some predicates hold, and can involve a 
set of state modifications. This section defines open automata and a bisimulation theory for them.

\subsection{Open Automata}
 Open automata (OA) are not composition structures but they are made of transitions that are dependent of the actions of the holes, and they can reason on variables (potentially with only symbolic values). Formaly, each state of an OA has set of \emph{state variables}, and all state variable sets are disjoint.
%\TODO{adopt a uniform notation for open transitions, almost each instance has a 
%different 
%notation! I suggest p,l,pr,po using \{\} for p,l,po as they are sets}
\begin{definition}[Open transitions]\label{def:OT}
	\label{def:OpenTransitions}
	An \emph{open transition} (OT) over a
	set $J$ of holes with sorts $\Sort_j^{j\in J}$ and a set of states $\mathcal{S}$ is 
	a structure of the form:	
	\begin{mathpar}
	\openrule
	{	\beta_j^{j\in J'}, \Pred, \Post}
	{s \OTarrow {\alpha}s'}
	\end{mathpar}
	Where $J'\subseteq J$, $s, s'\in\mathcal{S}$ and $\beta_j$
        is a transition of the hole $j$, with $\beta_j\in\Sort_j$. $\alpha$ is an action 
        label denoting the resulting action of this open transition.
        We call \emph{source variables} of an OT the union of all variables
        in the different terms
        $\beta_j$ and $\alpha$, and the state variables of $s$.
        \Pred\ is a predicate over the source variables. \Post\ is a set of 
	equations that hold \emph{after the open transition}, they are
        represented as a substitution of the form $({x_k\gets e_k})^{k\in K}$ 
	where $x_k$ are state variables of $s'$, and $e_k$ are expressions
        over the source variables. Open transitions are identified
        modulo logical equivalence on their predicate. 
\end{definition}

@@

\TODO{have a look at the end of the sec}

@@

\TODO{Eric: rewrite example with new Protocol}
\begin{example}\emph{An open-transition.}
  \label{OT:ABP}
At the upper level, the $SimpleSystem$ pNet has 2 holes and $SimpleProtocol$ as
a subnet. One of its possible open transitions (synchronizing the hole $P$
with the $Sender$ within the \emph{SimpleProtocol}) is:

@@

  The \emph{SimpleProtocol} pNet of Fig. \ref{SimpleProt:Spec} has 3 controllers and no holes. One of its possible open-transition (the sender sending $(m)$ to the medium) is:

 \smallskip\noindent
 $  OT_2  = \openrule{
   \emptyset; Pred=[send(SV2\_m)=send(s_1\_m);in(SV2\_m)=in(?m)];  \\ Post=\{m_1\_m:=?m\}
                      }
    {\ostate{s_1,m_0,r_0} \OTarrow{\underline{send(SV2\_m)}} \ostate{s_2,m_1,r_0}}
    $

    That can be simplified to:

    \ERIC{Apres application des substitutions et simplification du predicat:}

     $  OT_2  = \openrule{
      \emptyset; Pred=\True;  \\ Post=\{m_1\_m:=s_1\_m\}
                      }
    {\ostate{s_1,m_0,r_0} \OTarrow{\underline{send(s_1\_m)}} \ostate{s_2,m_1,r_0}}
    $

    \smallskip
    Note that the global states shows how the sender and the
    first medium have evolved during this transition, the sender
    moving from $s_1$ to $s_2$, and the first (forward) medium from
    $m_0$ to $m_1$.
    The predicate $Pred$ involves the state variables $s_1\_m$ and
    $s_1\_sb$ of the sender, while the assignment $Post$ sets the
    state variables of the medium.

@@

\paragraph{Variable management.}
The variables in each synchronisation vector are considered local:
for a given pNet expression, we must have fresh local variables for
each occurrence of a vector (i.e. each time we instantiate rule
\TrDeux). Similarly the state variables of each copy of a
given pLTS in the system, must be distinct, and those created for each
application of \TrDeux\ have to be fresh and all distinct. 
This is implemented within the open-automaton generation algorithm,
using name generation using a global counter as a suffix.

@@

Given an open-pNet
with finite synchronisation sets, finitely many leaves and
holes, and each pLTS at leaves having a finite number of states and
(symbolic) transitions, has a finite automaton (see \cite{henrio:Forte2016}). The algorithm for building such an automaton can be found in~\cite{QBMZ-AVOCS18}.

@@

\TODO{Eric: full (???) result of the tool execution on the running example on the Simple Prot Spec: Done, on the Counetr version !}

\TODO{:  and maybe one example on the Impl. OTs presented in latex readable form??}

@@

@@

\subsection{Bisimulation for open pNets -- a Composable Bisimulation Theory}
\label{section:bisimulation-PN}
As  our symbolic operational semantics provides an open automaton, we can apply the notion of
	strong (symbolic) bisimulation on automata to open pNets:
\begin{definition}[FH-bisimulation for open pNets]\label{def:bisim-pnets}
Two pNets are FH-bisimilar if there exist a relation between their associated 
automata that is an FH-bisimulation and their initial states are in the relation, i.e. 
the predicate associated to the relation between the initial states is \True.
\end{definition}
\TODO{Predicate=True me pose un probleme... typiquement, on a besoin d'une ``condition initiale'', pour avoir des pNets data-oriented equivalents a leur version state-oriebted}

@@

Finally, the previous theorems can be composed to state a general theorem about 
composability and FH-bisimilarity.
\begin{theorem}[Composability] \label{thm-composability}
	Consider two FH-bisimilar pNets with an arbitrary number of holes, when replacing, 
	inside those two original pNets, a subset of the holes by FH-bisimilar pNets, we 
	obtain two FH-bisimilar pNets.
\end{theorem}
This theorem is quite powerful somehow states that the theory of open pNets is convenient to study properties of process composition. Open pNets can indeed be used to study process operators and process algebras, as shown in~\cite{henrio:Forte2016}, or to study interaction protocols~\cite{BHHM:FACS11}.

@@

\TODO{Eric: try to draw (ATG) the weak transitino versions of the Simple Protocol, both Spec (easy) and Impl (check readability?)}

\subsection{Composition properties: composition of weak open transitions}
We now have two different semantics for open pNets: a stronf semantics, defined  as an open automaton, and as a weak semantics, defined as a weak open automaton. Like the open automaton, the weak open automaton features valuable composition properties. We can exhibit  a composition property and a decomposition property that relate open pNet composition with their semantics, defined as weak open automata. These are however technically more complex than the ones for open automata because each hole performs a set of actions, and thus a composed transition is the composition of one transition of the top-level pNet and a sequence of transitions of the sub-pNet that fills its hole. They can be found as Lemmas~\ref{lem-decomposeWOT}, Lemma~\ref{lem-Weakcompose1}, and Lemma~\ref{lem-Weakcompose} in~\ref{sec:app-composition}.

@@

\begin{definition}[Notation: Semantics of pNets as a weak open automaton]
Let $A$ be the open automaton expressing the semantics of an open \pNet: $\pNet$; let $<J,\mathcal{S},s_0,  \WT>$ be the weak open automata derived from $A$. For any $\WOT\in\WT$, we denote $\pNet \models \WOT$, this defines the semantics of the pNet $\pNet$ as a weak open automaton.
\end{definition}

@@

\TODO{Je ne pense pas que nous ayons un algo de bisim weak operationnel pour la publi, mais ca n'est pas completement impossible nonp[l;us...  Au pire, argumenter intuitivement sur papier, en donnat les classes d'equivalence et les predicats.
    Noter que la version de base est trop simple... Pour avoir des predicats moins triviaux, on peut ajouter un compteur qui compte le nb d'erreurs, et le transmet a la fin}

@@

\noindent
     $  \openrule{
      \emptyset; Pred=\True; Post=\{m_1\_m:=s_1\_m, m_1\_b:=s_1\_sb, s_2\_sb:=s_1\_sb\}
                      }
    {\ostate{s_1,m_0,m_0,r_0} \OTarrow{\underline{send(s_1\_m, s_1\_b)}} \ostate{s_2,m_1,m_0,r_0}}
    $

@@

@@

\begin{figure}[h]
  \centerline{\includegraphics[width=10cm]{ATG/SPImplOpen}}
  \caption{Open Automaton of the Simple Protocol Implementation}  \label{SimpleProtCounter:ImplOA}
\end{figure}

@@

>>>>

Fig. \ref{SimpleProt:Spec} (right) shows a graphical view of the pNet of the system \emph{specification}:
the full system should be equivalent (e.g. through weak bisimulation)
to the two processes connected simply through a perfect medium. \TODO{Decrire syntaxe graphique}

Fig. \ref{SimpleProt:Impl} shows the pNet model of the protocol implementation. When the $Medium$ detects an error (modeled by a local $\tau$ action), it sends back a $m\_error$ message to the $Sender$. The $Sender$ increments its local error counter $ec$, and resends the message (including $ec$) to the $Medium$, that will, eventually, transmit $m,ec$ to the $Receiver$.

@@

The semantics of open pNets will be defined  as an open automaton. An open
automaton is an automaton where each transition composes transitions of several LTSs with
action of some holes, the transition occurs if some predicates hold, and can involve a 
set of state modifications. This section defines open automata and a bisimulation theory for them.
\TODO{Add initialisation of state vars}

\subsection{Open Automata}
 Open automata (OA) are not composition structures but they are made of transitions that are dependent of the actions of the holes, and they can reason on variables (potentially with only symbolic values). Formaly, each state of an OA has a set of \emph{state variables}, and all state variable sets are disjoint.
%\TODO{adopt a uniform notation for open transitions, almost each instance has a 
%different 
%notation! I suggest p,l,pr,po using \{\} for p,l,po as they are sets}
\begin{definition}[Open transitions]\label{def:OT}
	\label{def:OpenTransitions}
	An \emph{open transition} (OT) over a
	set $J$ of holes with sorts $\Sort_j^{j\in J}$ and a set of states $\mathcal{S}$ is 
	a structure of the form:	
	\begin{mathpar}
	\openrule
	{	\beta_j^{j\in J'}, \Pred, \Post}
	{s \OTarrow {\alpha}s'}
	\end{mathpar}
	Where $J'\subseteq J$, $s, s'\in\mathcal{S}$ and $\beta_j$
        is a transition of the hole $j$, with $\beta_j\in\Sort_j$. $\alpha$ is an action 
        label denoting the resulting action of this open transition.
        We call \emph{source variables} of an OT the union of all variables
        in the different terms
        $\beta_j$ and $\alpha$, and the state variables of $s$.
        \Pred\ is a predicate over the source variables. \Post\ is a set of 
	equations that hold \emph{after the open transition}, they are
        represented as a substitution of the form $({x_k\gets e_k})^{k\in K}$ 
	where $x_k$ are state variables of $s'$, and $e_k$ are expressions
        over the source variables. Open transitions are identified
        modulo logical equivalence on their predicate. 
\end{definition}

@@

%\TODO{have a look at the end of the sec}

@@

\TODO{Eric: rewrite example with new Protocolkeep only OT1, and adapt comments}
\begin{example}\emph{An open-transition.}
  \label{OT:ABP}
At the upper level, the $SimpleSystem$ pNet has 2 holes and $SimpleProtocol$ as
a subnet. One of its possible open transitions (synchronizing the hole $P$
with the $Sender$ within the \emph{SimpleProtocol}) is:

@@

 %%  The \emph{SimpleProtocol} pNet of Fig. \ref{SimpleProt:Spec} has 3 controllers and no holes. One of its possible open-transition (the sender sending $(m)$ to the medium) is:

 %% \smallskip\noindent
 %% $  OT_2  = \openrule{
 %%   \emptyset; Pred=[send(SV2\_m)=send(s_1\_m);in(SV2\_m)=in(?m)];  \\ Post=\{m_1\_m:=?m\}
 %%                      }
 %%    {\ostate{s_1,m_0,r_0} \OTarrow{\underline{send(SV2\_m)}} \ostate{s_2,m_1,r_0}}
 %%    $

 %%    That can be simplified to:

 %%    \ERIC{Apres application des substitutions et simplification du predicat:}

 %%     $  OT_2  = \openrule{
 %%      \emptyset; Pred=\True;  \\ Post=\{m_1\_m:=s_1\_m\}
 %%                      }
 %%    {\ostate{s_1,m_0,r_0} \OTarrow{\underline{send(s_1\_m)}} \ostate{s_2,m_1,r_0}}
 %%    $

 %%    \smallskip
 %%    Note that the global states shows how the sender and the
 %%    first medium have evolved during this transition, the sender
 %%    moving from $s_1$ to $s_2$, and the first (forward) medium from
 %%    $m_0$ to $m_1$.
 %%    The predicate $Pred$ involves the state variables $s_1\_m$ and
 %%    $s_1\_sb$ of the sender, while the assignment $Post$ sets the
 %%    state variables of the medium.

@@

@@

We have shown in \cite{henrio:Forte2016} that an open-pNet
with finite synchronisation sets, finitely many leaves and
holes, and each pLTS at leaves having a finite number of states and
(symbolic) transitions, has a finite automaton. The algorithm for building such an automaton can be found in~\cite{QBMZ-AVOCS18}.

@@

%\TODO{Eric: full (???) result of the tool execution on the running example on the Simple Prot Spec: Done, on the Counetr version !}

\paragraph{Example} Figure \ref{SimpleProtCounter:SpecOA} shows the open automaton computed from the Specification pNet in Figure \ref{SimpleProt:Spec}.

%\TODO{:  and maybe one example on the Impl. OTs presented in latex readable form??}
\TODO{Comment on Fig 3...}

@@

Figure \ref{SimpleProtCounter:ImplOA} shows the full open automaton of the implementation $SimpleSystem$. Here, in the open automaton drawing, we use a user-friendly notation for the OTs, in the form ``$\{<actions of holes\} [<guard>] <action> \{assignments\}$''.
\TODO{comment on difference from Exemple 1 page 15; and predicate simplification...}
\begin{figure}[h]
  \centerline{\includegraphics[width=10cm]{ATG/SPImplOpen}}
  \caption{Open Automaton of the Simple Protocol Implementation}  \label{SimpleProtCounter:ImplOA}
\end{figure}

%% We show here the first open transition, written here in the form of the tool full output.

%% $  OT_1  = \openrule{
%%   (P\mapsto p\_send(m:sva\_SV6:1:1)) \\
%%   [( msg =m: sva\_SV6 :1:1) \wedge (( in( msg ))=( in(m: sva_SV6 :1:1)))]\\
%%   \{ s1\_msg := msg , s1\_c := 0\}
%%                       }
%% {\ostate{s_0,m_0,r_0} \OTarrow{\underline{in(m:sva\_SV6:1:1)}} \ostate{s_1,m_0,r_0}}
%% $
    \TODO{Comment...}

@@

\subsection{Bisimulation for open pNets -- a Composable Bisimulation Theory}
\label{section:bisimulation-PN}
As  our symbolic operational semantics provides an open automaton, we can apply the notion of
	strong (symbolic) bisimulation on automata to open pNets:
\begin{definition}[FH-bisimulation for open pNets]\label{def:bisim-pnets}
Two pNets are FH-bisimilar if there exist a relation between their associated 
automata that is an FH-bisimulation and their initial states are in the relation (with some initial predicate).
\TODO{check}
\end{definition}
%\TODO{Predicate=True me pose un probleme... typiquement, on a besoin d'une ``condition initiale'', pour avoir des pNets data-oriented equivalents a leur version state-oriebted => remove the constraint}

@@

Finally, the previous theorems can be composed to state a general theorem about 
composability and FH-bisimilarity.
\begin{theorem}[Composability] \label{thm-composability}
	Consider two FH-bisimilar pNets with an arbitrary number of holes, when replacing, 
	inside those two original pNets, a subset of the holes by FH-bisimilar pNets, we 
	obtain two FH-bisimilar pNets.
\end{theorem}
This theorem is quite powerful. It somehow implies that the theory of open pNets is convenient to study properties of process composition. Open pNets can indeed be used to study process operators and process algebras, as shown in~\cite{henrio:Forte2016}, or to study interaction protocols~\cite{BHHM:FACS11}.

@@

\begin{example}
  Figure \ref{SimpleProtCounter:SpecOA-partial} show a partial view of the weak automaton of the simple protocol implementation.

\TODO{Explain the partial OA, and comment on building of WOT1}
\end{example}

\begin{figure}[t]
   \centerline{\includegraphics[width=12cm]{ATG/SPImplWeakOpen}}
  \caption{Weak Open Automaton, partial}
   \label{SimpleProtCounter:SpecOA-partial}
\end{figure}

\subsection{Composition properties: composition of weak open transitions}
We now have two different semantics for open pNets: a strong semantics, defined  as an open automaton, and as a weak semantics, defined as a weak open automaton. Like the open automaton, the weak open automaton features valuable composition properties. We can exhibit  a composition property and a decomposition property that relate open pNet composition with their semantics, defined as weak open automata. These are however technically more complex than the ones for open automata because each hole performs a set of actions, and thus a composed transition is the composition of one transition of the top-level pNet and a sequence of transitions of the sub-pNet that fills its hole. They can be found as Lemmas~\ref{lem-decomposeWOT}, Lemma~\ref{lem-Weakcompose1}, and Lemma~\ref{lem-Weakcompose} in~\ref{sec:app-composition}.

@@

\begin{definition}[Notation: Semantics of pNets as a weak open automaton]
Let $A$ be the open automaton expressing the semantics of an open pNet $\pNet$; let $<J,\mathcal{S},s_0,  \WT>$ be the weak open automata derived from $A$. For any $\WOT\in\WT$, we denote $\pNet \models \WOT$, this defines the semantics of the pNet $\pNet$ as a weak open automaton. \TODO{strange sentence}
\end{definition}

@@

\TODO{Je ne pense pas que nous ayons un algo de bisim weak operationnel pour la publi, mais ca n'est pas completement impossible nonp[l;us...  Au pire, argumenter intuitivement sur papier, en donnat les classes d'equivalence et les predicats.}

@@

\TODO{Eric: rewrite completely}
\noindent
     $  \openrule{
      \emptyset; Pred=\True; Post=\{m_1\_m:=s_1\_m, m_1\_b:=s_1\_sb, s_2\_sb:=s_1\_sb\}
                      }
    {\ostate{s_1,m_0,m_0,r_0} \OTarrow{\underline{send(s_1\_m, s_1\_b)}} \ostate{s_2,m_1,m_0,r_0}}
    $

@@

\TODO{keep ?}
\paragraph{Variable management.}
The variables in each synchronisation vector are considered local:
for a given pNet expression, we must have fresh local variables for
each occurrence of a vector (i.e. each time we instantiate rule
\TrDeux). Similarly the state variables of each copy of a
given pLTS in the system, must be distinct, and those created for each
application of \TrDeux\ have to be fresh and all distinct. 
This is implemented within the open-automaton generation algorithm,
using name generation using a global counter as a suffix.

@@

\begin{figure}[h]
  \centerline{\includegraphics[width=10cm]{ATG/SPImplOpen}}
  \caption{Open Automaton of the Simple Protocol Implementation}  \label{Annex:SimpleProtCounter:ImplOA}
\end{figure}

@@

In Figure \ref{Annex:SimpleProtCounter:ImplWOA}, we have added some of the weak transitions generated by the saturation, namely:
- a tau-loop on each state
- a transition from <s0\_a0\_r0> to <s2\_a1\_r0>, resulting from the sequence OT\-1;OT\-6 followed by any number of the loop (OT\_9;ot\_12;OT\_6).

Many weak should be added, e.g. Tau loops on states <s2\_a1\_r0>, <s2\_a2\_r0>, 
<s2\_a0\_r0>, encoding the loop.
